---
title: "Simple autoregressive (AR) and Generalized Additive Model (GAM) in JAGS"
author: "Petr Keil"
date: "January 22, 2016"
output: 
  html_document: 
    highlight: pygments
    number_sections: yes
    theme: cerulean
    toc: yes
---

The data are the [HadCRUT3v](https://goo.gl/18n6FB) data set, which I actually took from [this blogpost](http://goo.gl/OIHMhC). Here I reworked the analysis for the Bayesian setting. The data decribe **mean global annual temperature anomalies from 1850 to 2013**, provided at yearly intervals.

# The data

```{r, fig.width=12}
   gtemp <- read.csv("http://www.petrkeil.com/wp-content/uploads/2016/01/gtemp.csv")
   plot(Annual~Year, data=gtemp, type="b")
```

Let's explore the **temporal autocorrelation** a bit:

```{r, fig.width=12}
   par(mfrow=c(1,2))
   plot(gtemp$Annual[-164], gtemp$Annual[-1],
        xlab="T in Year", ylab="T in Year-1")
   acf(gtemp$Annual, main="")
```

# First-order autoregressive (AR1) model in JAGS

Formal definition of the 1st-order autoregressive (AR1) model:

$$y_t = \beta_0 + \beta_1 y_{t-1} + \epsilon_i$$

## Prepare the data

```{r}
   ar.data <- list(Annual = gtemp$Annual,
                   N = nrow(gtemp))
   
```

## The model

We will use the `R2jags` library:

```{r, message=FALSE}
   library(R2jags)
```

JAGS model specification:

```{r}
cat("
  model
  {
    # priors
    tau <- 1/(sigma*sigma)
    sigma ~ dunif(0, 100)      
    beta0 ~ dnorm(0, 0.001)
    beta1 ~ dnorm(0, 0.001)
    mu[1] ~ dnorm(0, 0.001)

    # likelihood
    eps[1] <- Annual[1] - mu[1]

    for(t in 2:N)
    {
      mu[t] <- beta0 + beta1*Annual[t-1]
      eps[t] <- Annual[t] - mu[t]
    }  
    for(t in 1:N)
    {
      Annual[t] ~ dnorm(mu[t], tau)
    }
  }
", file="ar1.txt")
```

Fitting the model:

```{r, message=FALSE}   
   model.fit <- jags(data=ar.data, 
               model.file="ar1.txt",
               parameters.to.save=c("beta0", "beta1"),
               n.chains=3,
               n.iter=2000,
               n.burnin=1000)
   model.fit
```

```{r, message=FALSE, fig.width=12}   
   mus <- jags(data=ar.data, 
               model.file="ar1.txt",
               parameters.to.save=c("mu"),
               n.chains=3,
               n.iter=2000,
               n.burnin=1000, 
               DIC=FALSE)
   mus <- data.frame(mus$BUGSoutput$summary)
   
   plot(Annual~Year, data=gtemp, type="b")
   lines(gtemp$Year, mus$X50., col="red", lwd=2)
```

# First order noving Average (MA1) model

JAGS model specification:

```{r}
cat("
  model
  {
    # priors
    tau <- 1/(sigma*sigma)
    sigma ~ dunif(0, 100)      
    beta0 ~ dnorm(0, 0.001)
    beta1 ~ dnorm(0, 0.001)
    mu[1] ~ dnorm(0, 0.001)

    # likelihood
    
    eps[1] <- Annual[1] - mu[1]

    for(t in 2:N)
    {
      mu[t] <- beta0 + beta1*eps[t-1]
      eps[t] <- Annual[t] - mu[t]
    }  
    for(t in 1:N)
    {
      Annual[t] ~ dnorm(mu[t], tau)
    }
  }
", file="ma1.txt")
```

```{r, message=FALSE, fig.width=12}   
   mus <- jags(data=ar.data, 
               model.file="ma1.txt",
               parameters.to.save=c("mu"),
               n.chains=3,
               n.iter=2000,
               n.burnin=1000, 
               DIC=FALSE)
   mus <- data.frame(mus$BUGSoutput$summary)
   
   plot(Annual~Year, data=gtemp, type="b")
   lines(gtemp$Year, mus$X50., col="red", lwd=2)
```

# ARMA[1,1] model

```{r}
cat("
  model
  {
    # priors
    tau <- 1/(sigma*sigma)
    sigma ~ dunif(0, 100)      
    beta0 ~ dnorm(0, 0.001)
    beta1 ~ dnorm(0, 0.001)
    beta2 ~ dnorm(0, 0.001)
    mu[1] ~ dnorm(0, 0.001)

    # likelihood
    
    eps[1] <- Annual[1] - mu[1]

    for(t in 2:N)
    {
      mu[t] <- beta0 + beta1*eps[t-1] + beta2*mu[t-1]
      eps[t] <- Annual[t] - mu[t]
    }  
    for(t in 1:N)
    {
      Annual[t] ~ dnorm(mu[t], tau)
    }
  }
", file="arma.txt")
```

```{r, message=FALSE, fig.width=12}   
   mus <- jags(data=ar.data, 
               model.file="arma.txt",
               parameters.to.save=c("mu"),
               n.chains=3,
               n.iter=2000,
               n.burnin=1000, 
               DIC=FALSE)
   mus <- data.frame(mus$BUGSoutput$summary)
   
   plot(Annual~Year, data=gtemp, type="b")
   lines(gtemp$Year, mus$X50., col="red", lwd=2)
```


# Generalized additive model (GAM) model in R

```{r, fig.width=12}
   library(mgcv)

   m1 <- gam(Annual ~ s(Year), data=gtemp)

   m1.preds <- predict(m1, type="response", se.fit=TRUE)
   
   plot(Annual~Year, data=gtemp, type="b")
      lines(gtemp$Year, m1.preds$fit, lwd=2, col="red")
      lines(gtemp$Year, m1.preds$fit-m1.preds$se.fit, col="red", lty=2)
      lines(gtemp$Year, m1.preds$fit+m1.preds$se.fit, col="red", lty=2)
```

# Fit the GAM model in JAGS

# Use the `jagam` function to build the model

```{r}
    jagam.model <- jagam(Annual ~ s(Year), 
                         data=gtemp, 
                         file="jagam_model.txt")
    readLines("jagam_model.txt")
```

# Fit the model
```{r}
   library(R2jags)
   model.fit <- jags(data=jagam.model$jags.data, 
                     model.file="jagam_model.txt",
                     parameters.to.save=c("mu"),
                     n.chains=3,
                     n.iter=1000,
                     n.burnin=500,
                     DIC=FALSE)
```

# Plot the predicted mean

```{r, fig.width=12}
   preds <- data.frame(model.fit$BUGSoutput$summary)
   plot(Annual~Year, data=gtemp, type="b", main="Bayesian estimate of the mean")
     lines(gtemp$Year, preds$X50., col="red", lwd=2)
     lines(gtemp$Year, preds$X2.5., lty=2, col="red")
     lines(gtemp$Year, preds$X97.5., lty=2, col="red")
     
  acf(gtemp$Annual-preds$X50.)
```

# Plot the full prediction interval

```{r}
cat("
model {
  mu <- X %*% b ## expected response
  for (i in 1:n) 
  { 
     y[i] ~ dnorm(mu[i],tau) 
  } ## response 
  
  scale <- 1/tau ## convert tau to standard GLM scale
  tau ~ dgamma(.05,.005) ## precision parameter prior 
  ## Parametric effect priors CHECK tau=1/1.7^2 is appropriate!
  for (i in 1:1) { b[i] ~ dnorm(0,0.36) }
  ## prior for s(Year)... 
  K1 <- S1[1:9,1:9] * lambda[1]  + S1[1:9,10:18] * lambda[2]
  b[2:10] ~ dmnorm(zero[2:10],K1) 
  ## smoothing parameter priors CHECK...
  for (i in 1:2) {
    lambda[i] ~ dgamma(.05,.005)
    rho[i] <- log(lambda[i])
  }

  # PREDICTIONS
  for (i in 1:n)
  {
     pred[i] ~ dnorm(mu[i],tau) 
  }
}
", file="jagam_model_with_preds.txt")
```

# Fit the model
```{r}
   library(R2jags)
   model.fit <- jags(data=jagam.model$jags.data, 
                     model.file="jagam_model_with_preds.txt",
                     parameters.to.save=c("pred"),
                     n.chains=3,
                     n.iter=1000,
                     n.burnin=500,
                     DIC=FALSE)
```

# Plot the predicted mean

```{r, fig.width=12}
   preds <- data.frame(model.fit$BUGSoutput$summary)
   plot(Annual~Year, data=gtemp, type="b", main="Full Bayesian prediction interval")
     lines(gtemp$Year, preds$X50., col="red", lwd=2)
     lines(gtemp$Year, preds$X2.5., lty=2, col="red")
     lines(gtemp$Year, preds$X97.5., lty=2, col="red")
```

# Combining the AR1 and GAM model

This section is yet unfinished.





